Optimal Finite Element Integration

### 2015-08-12 TOMS_Paper/assembly.tex
Refine second half of paper, update elasticity plot

For what concerns the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L2 cache (the last level of private cache). For the test cases in which $T_H$ was determinant to prevent pre-evaluation, we also repeated the experiments setting $T_H = L3_{size}$ and $T_H = \infty$. Results of these trials are discussed below.

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix .}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model}\label{fig:hyperelasticity}
\end{figure}

@@

\paragraph{Mass}
From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} represents the best variant among the ones tested, apart from the few points previously discussed. This suggests that the optimality model holds in these simple bilinear forms; that is, \texttt{cfO2} generates loop nests that are optimal from both a theoretical and practical viewpoints. It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in many of the test cases, especially as $nf \geq 2$.

\paragraph{Helmholtz}
Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant. We appreciate the fact that the relative order of the employed function spaces plays a key role in the code synthesis choice. Take, for example, the case $[q=3, p=1, nf=2]$. The adoption of pre-evaluation makes a tremendous speed up be achievable by $cfO2$ over the competitors. On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal.

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which \texttt{tens} (and, generally, pre-evaluation through \texttt{cfO2}), tends to be the best alternative when a single coefficient is present. With $q=4$ and $nf \in [0, 1]$, pre-evaluation in \texttt{cfO2} is found profitable in terms of operation count, although it is not triggered because of exceeding the working set threshold $T_H$. Running the same experiments with $T_H = \infty$ resulted in significant runtime increases for all values of $p$. Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement for $nf = 0$, with \texttt{cfO2} becoming the fastest alternative. Our explanation for this is that by running a sequential program, a process has exclusive access to the memory bandwidth of the system, so pre-fetching minimizes cache misses without hurting concurrent processes (since there are none).

@@

\section{Conclusions}
\label{sec:conclusions}
With this research we have made a first step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration. The results are extremely encouraging, suggesting our model applies to a variety of contexts. We have discussed the conditions under which the optimality model becomes weaker. An open problem is understanding how to refine this model to include outer loops. This will probably require exploiting properties of differential operators. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces.

>>>>

For what concerns the memory constraint $C_2$ introduced in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L2 cache (the last level of private cache). For the test cases in which $T_H$ was determinant to prevent pre-evaluation, we also repeated the experiments setting $T_H = L3_{size}$ to assess our hypotheses. Results of these trials are discussed below.

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

@@

\paragraph{Mass}
From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} is the best variant among the ones tested, apart from a few points already discussed. This suggests that the optimality model holds in these simple bilinear forms; that is, \texttt{cfO2} generates loop nests that are optimal from both a theoretical and practical viewpoints. It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in many of the test cases, especially as $nf \geq 2$. The fact that \texttt{cfO2} is slower than \texttt{tens} at $[q=4, p=4, nf=1]$ is explained by the memory threshold that prevents pre-evaluation. By just setting $T_H = L3_{size}$, \texttt{cfO2} becomes 1.48$\times$ faster than \texttt{tens}. This is probably due to the hardware prefetching being way more effective than in all other forms: the number of allocated temporaries is relatively small, and so is the number of memory access traces that need be tracked. However, as we already explained, in a parallel context this gain might diminish, since the L3 cache is shared by the cores of the CPU.

\paragraph{Helmholtz}
Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant. We appreciate the fact that the relative order of the employed function spaces plays a key role in the code synthesis choice. Take, for example, the case $[q=3, p=1, nf=2]$. The adoption of pre-evaluation makes a tremendous speed up be achievable by \texttt{cfO2} over the competitors. On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal. The effects of the working set threshold are in general positive. Moving to $T_H = L3_{size}$ makes \texttt{cfO2} slower than \texttt{cfO1} in many cases with $nf=2$, the most significant one being a slow down of 2.16$\times$ at $[q=2, p=2]$.

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which the choice of \texttt{tens} was generally correct when $nf=1$. Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$. In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to completely removing sharing and avoiding iteration over zero-valued blocks. The latter case occurs with $q=4$ and $nf \in [0, 1]$. Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow downs characterized the cases of $nf=1$ for all values of $p$. Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only roughly 40$\%$, while the cost function predicts a save in operation count of 5$\times$. This suggest that our model could be refined to handle the cases in which the gain in operation count is so large to make non-extreme variations in working set size negligible.

@@

\section{Conclusions}
\label{sec:conclusions}
With this research we have made a first step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration. The results are extremely encouraging, suggesting our model applies to a variety of contexts. We have discussed the conditions under which the optimality model becomes weaker. An open problem is understanding how to refine this model to include outer loops. This will probably require exploiting properties of differential operators. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces. Lastly, we recall our work is publicly available and is already in use in the latest version of the Firedrake framework.
